\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-13 k-planar.tex
Added figure

% \note{DW}{I suggest we use consecutive numbering of theorems, lemmas, etc. Theorem 1, Lemma 2, Observation 3, etc.}

@@

\subsection{\boldmath  $1$-Planar Graphs}
\label{sec-1-planar}

@@

Let $G$ be a $1$-plane graph.  We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

Consider some pair of edges $vw$ and $xy$ that cross at $p$.  The point $p$ is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$ and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three or four edges; see \cref{one_planar_example}(b).  Since, for any pair of crossing edges $\{vw,xy\}$ that we remove this way, $vxwy$ is a cycle in $G'$, $\dist_{G'}(v,w)\le 2$ and $\dist_{G'}(x,y)\le 2$.
% It follows immediately that, for any pair of vertices $a,b\in V(G)$, $\dist_{G'}(a,b)\le 2\dist_G(a,b)$.

A tree-decomposition $\mathcal{T}$ of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $3$-element subset $\{u,v,w\}\subseteq V(G)$, at most two bags of $\mathcal{T}$ contain $\{u,v,w\}$.  Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $|V(P_i)\cap L_j| \le 3$ for every integer $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
        \item For each $x\in V(H)$, $S_x=Y_x\cup X_x$ where $|X_x|\le 1$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
		\item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$..
	\end{compactenum}
\end{lem}

\begin{proof}
  This proof is very similar to the proof of Lemma~13 by \citet{DJMMUW20}, but is sufficiently complicated by the $1$-planar setting to warrant a full proof.  In particular, the reader should keep in mind that $G$ and $G'$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

  The proof is by induction on the number of inner faces of $N'$.  A particular easy case occurs when $V(N')=V(F)$, for which we can simply take $\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iv):
  \begin{compactenum}[(i)]
    \item By definition, each of $P_1,\ldots,P_k$ is a part of $\mathcal{P}$.
    \item For each $x\in V(H)$, let $Y_x:=S_x$ and $X_x=\emptyset$.  Then 4(a) ensures that $|Y_x\cap L_j|\le 3$ for each integer $j\ge 0$.
    \item The trivial tree decomposition of $H$ that places $x_1,\ldots,x_k$ in one bag has width $k-1\le 2\le 3$.
  \end{compactenum}

@@

    \item If $k=3$ then assign each vertex $v$ of $N'$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $F$ encountered on the path in $T$ from $v$ to the root of $T$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now add one edge to each $4$-sided inner face of $N'$ to obtain a plane supergraph $N''$ of $N'$ in which each inner face has exactly three edges on its boundary. By Sperner's Lemma there exists an inner face $\tau'':=v_1,v_2,v_3$ of $N''$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.  Let $\tau:=v_1,\ldots,v_p$ be the face of $N'$ that contains $\tau''$.  % (Note that $p\in\{3,4\}$.)
  \end{compactenum}

@@

  We can now define the vertices in $S$.  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to a vertex in $V(F)$.  Let $\overline{S}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $S:=V(\overline{S})\setminus V(F)$, let $Y:=(V(Q_1)\cup V(Q_2)\cup V(Q_3))\setminus V(F)$, and let $X:=S\setminus Y$.
  Observe that, for each $i\in\{1,2,3\}$ and each $j\ge 0$, $|V(Q_i)\cap L_j|\le 1$.  Therefore $|Y\cap L_j|\le 3$ for each integer $j\ge 0$. Furthermore, $|X|\le p-3\le 1$, since $v_1,v_2,v_3\in Y\cup V(F)$. Therefore, using $S:=Y\cup X$ as a part in the partition $\PP$ will not violate condition~(ii).

@@

  \begin{figure}
    \begin{center}
      \begin{tabular}{ccc}
        \includegraphics{figs/mmplus-1} &
        \includegraphics{figs/mmplus-2} &
        \includegraphics{figs/mmplus-3} \\
        $k=1$ & $k=2$ & $k=3$
      \end{tabular}
    \end{center}
    \caption{Three cases when choosing the face $\tau$ of $N'$.}
    \label{boring_figure}
  \end{figure}

@@

We remark that our proof of \cref{1-planar} generalizes to the class of $d$-framed graphs \cite{BDGGMR} and even $d$-framed multigraphs.  An embedded multigraph $G$ is a \defin{$d$-framed multigraph} if it contains a plane subgraph $G'$ (the \defin{frame}) whose edges are not involved in any crossings and such that each pair of crossing edges in $G$ are contained in a face of $G'$ that has at most $d$ vertices on its boundary.  It is known that every $3$-connected $1$-planar graph is contained in some $4$-framed graph.  The discussion on the structure of $G$ and $G'$ that precedes \cref{induction} shows that every $1$-planar graph is contained in a $4$-framed multigraph.
The proof of \cref{induction} can be repeated for any edge-maximal $d$-framed multigraphs $G$ having frame $G'$. The proof in this case has only two differences:
\begin{compactitem}
   \item For any edge $vw$ of $G$ that is not in $G'$, $\dist_{G'}(v,w)\le \lfloor d/2\rfloor$.
   \item The face $\tau$ that contains the Sperner triangle $v_1,v_2,v_3$ may have up to $d$ vertices on its boundary, so up to $d-3$ of these vertices do not belong to $Q_i$ for any $i\in\{1,2,3\}$.
\end{compactitem}
The only effect this has on the final result is that the layered width becomes $3\lfloor d/2\rfloor + d-3$.  This yields the following result:

\begin{thm}
\label{framed}
   Every $d$-framed multigraph is contained in $H\boxtimes P\boxtimes K_{3\lfloor d/2\rfloor + d -3}$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

\note{DW}{I think \cref{framed} should be proved explicitly and \cref{1-planar} should be a corollary, especially because \cref{NewMapGraph} is another corollary. The proof of \cref{framed} needs an extra clause; see the last sentence of \cref{NewLemma}. This section should be about framed graphs, outside of \cref{sec-k-planar}. I suggest we make it a subsection of \cref{Structure} where we introduce framed graphs, say they can be described by $(2,\binom{d}{2})$-shortcut systems, so the main theorem is applicable, which gives a treewidth bound of $\binom{5}{3}-1=9$, but using  \cref{framed} we get a treewidth bound of 3, which is optional. Then in latter sections we give two applications, for 1-planar graphs and for $d$-map graphs. }

\note{PM}{I no longer agree about the restructuring. See my email essay for an explanation.}

\note{PM}{Apropos the extra clause.  It's related to the orange todo note (argue that $N'/\mathcal{P}=N/\mathcal{P}$).  This doesn't need to be in the statement of the lemma since, unlike \cref{NewLemma} this one gives an $H$-partition of $N$, not $N'$, but something like it needs to be in the proof.}

>>>>

\note{PM}{The trick I used (using the thm counter everywhere) to get consecutive Theorem, Lemma, etc numbering has broken cleverref.  Everything is now references as a Theorem.  Find a solution that works.}
% \note{DW}{I suggest we use consecutive numbering of theorems, lemmas, etc. Theorem 1, Lemma 2, Observation 3, etc.}

@@

\subsection{\boldmath $d$-Framed and $1$-Planar Graphs}
\label{sec-1-planar}

@@

We will actually prove this result in the more general setting of $d$-framed graphs \cite{BDGGMR}. For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it admits a drawing having a biconnected plane skeleton (multi)graph $G_0$, whose faces have size at most $d$, and whose crossing edges are in the interior of these faces.  In \cref{1_planar_is_4_framed}, below, we show that every $1$-planar graph is a subgraph of a $4$-framed multigraph. This makes \cref{1-planar} an immediate consequence of the following theorem, whose proof is the subject of the rest of this section.

\begin{thm}\label{d_framed_product_stucture}
  Every $d$-framed graph is contained in $H\boxtimes P\boxtimes K_{\lfloor d/2\rfloor + d - 3}$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

\subsection{\boldmath $d$-Framed Graphs}

To establish the planarity of $H$ in \cref{d_framed_product_stucture}, the following definition is helpful.
A tree-decomposition $\mathcal{T}$ of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $3$-element subset $\{u,v,w\}\subseteq V(G)$, at most two bags of $\mathcal{T}$ contain $\{u,v,w\}$.  Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ be a $d$-framed graph and let $G'$ be the frame of $G$.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $P_i=Y_i\cup X_i$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
        \item For each $x\in V(H)$, $S_x=Y_x\cup X_x$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
		\item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
	\end{compactenum}
\end{lem}

\begin{proof}
  This proof is very similar to the proof of Lemma~13 by \citet{DJMMUW20}, but is complicated by several factors.  In particular, the reader should keep in mind that $G$ and $G'$ are multigraphs and that the cycle $F$ may consist of two vertices and two (parallel) edges.

  The proof is by induction on the number of inner faces of $N'$.  A particular easy case occurs when $V(N')=V(F)$, for which we can simply take $\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iv):
  \begin{compactenum}[(i)]
    \item By definition, each of $P_1,\ldots,P_k$ is a part of $\mathcal{P}$.
    \item Assumption 4(a) ensures, for each $x\in V(H)$, the existence of $X_x$ and $Y_x$ such that $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each integer $j\ge 0$.
    \item The trivial tree decomposition of $H$ that places $x_1,\ldots,x_k$ is $3$-simple and, if $k=3$, has exactly one bag that contains $\{x_1,\ldots,x_3\}$.
  \end{compactenum}

@@

    \item If $k=3$ then assign each vertex $v$ of $N'$ a colour $\alpha(v)\in\{1,\ldots,3\}$ as follows:  Let $w$ be the first vertex of $F$ encountered on the path in $T$ from $v$ to the root of $T$.  (This implies that $w=v$ when $v$ is a vertex of $F$.)  Since $w\in V(F)$, $w\in P_i$ for some $i\in\{1,\ldots,k\}$ and we define $\alpha(v):=i$.  Now, for each $d'\in\{4,\ldots,d\}$ and each $d'$-sided inner face $F$ of $N'$, add edges $d'-2$ edges inside of $F$ to split it into $3$-sided faces.  This yields a plane supergraph $N''$ of $N'$ in which each inner face has exactly three edges on its boundary. By Sperner's Lemma there exists an inner face $\tau'':=v_1,v_2,v_3$ of $N''$ such that $\alpha(v_i)=i$ for each $i\in\{1,2,3\}$.  Let $\tau:=v_1,\ldots,v_p$ be the face of $N'$ that contains $\tau''$.  % (Note that $p\in\{3,4\}$.)
  \end{compactenum}

@@

  We can now define the vertices in $S$.  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to a vertex in $V(F)$.  Let $\overline{S}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $S:=V(\overline{S})\setminus V(F)$, let $Y:=(V(Q_1)\cup V(Q_2)\cup V(Q_3))\setminus V(F)$, and let $X:=S\setminus Y$.
  Observe that, for each $i\in\{1,2,3\}$ and each $j\ge 0$, $|V(Q_i)\cap L_j|\le 1$.  Therefore $|Y\cap L_j|\le 3$ for each integer $j\ge 0$. Furthermore, $|X|\le p-3\le d-3$, since $v_1,v_2,v_3\in Y\cup V(F)$. Therefore, using $S:=Y\cup X$ as a part in the partition $\PP$ will not violate condition~(ii).

@@

  \begin{figure}
    \begin{center}
      \begin{tabular}{ccc}
        \includegraphics{figs/mmplus-1} &
        \includegraphics{figs/mmplus-2} &
        \includegraphics{figs/mmplus-3} \\
        $k=1$ & $k=2$ & $k=3$
      \end{tabular}
    \end{center}
    \caption{Three cases when choosing the face $\tau$ of $N'$.}
    \label{boring_figure}
  \end{figure}
  \todo[inline]{Update these figures with $d$-sided $\tau$.}

@@

\subsection{\boldmath $1$-Planar Graphs}

\begin{lem}\label{1_planar_is_4_framed}
  A graph $G$ is $1$-planar if and only if it is contained in a $4$-framed multigraph.
\end{lem}

\begin{proof}
Let $G$ be a $1$-plane graph.  We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

Consider some pair of edges $vw$ and $xy$ that cross at $p$.  The point $p$ is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$ and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three or four edges; see \cref{one_planar_example}(b).  Since, for any pair of crossing edges $\{vw,xy\}$ that we remove this way, $vxwy$ is a cycle in $G'$, $\dist_{G'}(v,w)\le 2$ and $\dist_{G'}(x,y)\le 2$.
% It follows immediately that, for any pair of vertices $a,b\in V(G)$, $\dist_{G'}(a,b)\le 2\dist_G(a,b)$.
\end{proof}
